---
title: "Notes20251117"
author: "Blazej Kochanski"
date: "2025-11-17"
output: html_document
---

We will use ggplot2 package. 

```{r}
library(ggplot2)
```

UsingR packages has the Pearson's data from ca. 1900. 

```{r}
#install.packages("UsingR")
```

```{r}
library(UsingR)
data(father.son)
head(father.son)
```

Now let us build the scatter plot of fathers and sons' height. Father's height on x-axis and son's height on y axis. 

```{r}
ggplot(data = father.son, 
       mapping = aes(x = fheight, y = sheight)) +
  geom_point()
```


Add `alpha = 0.4` to see through the points (they will be 60% transparent). 

```{r}
ggplot(data = father.son, 
       mapping = aes(x = fheight, y = sheight)) +
  geom_point(alpha = 0.4)
```

The descriptions of the axes. 

```{r}
ggplot(data = father.son, 
       mapping = aes(x = fheight, y = sheight)) +
  geom_point(alpha = 0.4, col="blue") +
  xlab("Father's height [inches]") +
  ylab("Son's height [inches]") +
  ggtitle("Scatter plot of Pearson's father/son data")
```


Let us colour the dots where fathers where ~70 inches tall. 

```{r}
# Auxiliary table with the families where rounded height of the father is 70 inches
df70 <- father.son[round(father.son$fheight)==70, ]
```

```{r}
ggplot(data = father.son, 
       mapping = aes(x = fheight, y = sheight)) +
  geom_point(alpha = 0.4, col="black") +
  geom_point(data = df70, alpha = 0.4, col = "blue") +
  xlab("Father's height [inches]") +
  ylab("Son's height [inches]") +
  ggtitle("Scatter plot of Pearson's father/son data")
```

Now, colour the dots where fathers where ~67 inches tall (in green) and ~64 inches tall (in yellow). 

```{r}
# Auxiliary tables
df67 <- father.son[round(father.son$fheight)==67, ]
df64 <- father.son[round(father.son$fheight)==64, ]

G <- ggplot(data = father.son, 
       mapping = aes(x = fheight, y = sheight)) +
  geom_point(alpha = 0.4, col="black") +
  geom_point(data = df70, alpha = 0.4, col = "blue") +
  geom_point(data = df67, alpha = 0.4, col = "chartreuse4") +
  geom_point(data = df64, alpha = 0.4, col = "yellow2") +
  xlab("Father's height [inches]") +
  ylab("Son's height [inches]") +
  ggtitle("Scatter plot of Pearson's father/son data")

G 
```

