---
title: "Notes20251201"
author: "Blazej Kochanski"
date: "2025-12-15"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

Principal component analysis: https://en.wikipedia.org/wiki/Principal_component_analysis

We will use the data on fake and random series of coin tosses:

link: https://raw.githubusercontent.com/blakocha/ADRPY2025/refs/heads/main/cointosses.csv

```{r}
url = "https://raw.githubusercontent.com/blakocha/ADRPY2025/refs/heads/main/cointosses.csv"
coin <- read.csv(url) 
head(coin)
```

The file cointains the data on `r sum(coin$fake)` fake series of 100 coin tosses each generated by students on paper, as well as `r sum(!coin$fake)` random series of 100 coin tosses generated with the help of the (pseudo-)random number generator in R. 

Variables: 

- **toss** - a series of 100 coin toss results coded as O ("orzel", _tails_ in Polish) and R ("reszka", _heads_ in Polish), 
- **fake** - an indicator variable, 1 if fake, 0 if randomly generated, 
 
- **pierwszy_O** - pre-computed variable indicating whether the first toss in the series is "O"
 
- **liczba_serii** - number of runs (for example, in series: "OOORROOR" there are 4 runs: OOO, RR, OO, R)
 
- **liczba_O** - number of Os in the run
 
- **max_seria** - length of the longest run (for example, if the longest run is "OOOOO", max_seria is 5)
 
- **runtest_z** - test statistic from the [Wald-Wolfowitz runs test](https://en.wikipedia.org/wiki/Wald%E2%80%93Wolfowitz_runs_test) 
 
- **sd10** - standard deviation of number of Os in each row of ten tosses
 
- **sum_2max**, **sum_3max**, **sum_4max**, etc. - sum of lengths of the 2, 3, 4, etc. longest series
 
- **count_1** - number of runs of length 1

```{r}
library(ggplot2)

ggplot(coin, aes(x = max_seria, y = liczba_serii, col=as.factor(fake))) +
         geom_point(alpha = 0.5)

```

```{r}
library(plotly)

plot_ly(
  data = coin,
  x = ~max_seria,
  y = ~liczba_serii,
  z = ~liczba_O,
  color = ~as.factor(fake),
  type = "scatter3d",
  mode = "markers",
  marker = list(opacity = 0.5)
)
```

```{r}
X <- coin[,4:16]
pca <- prcomp(X, scale. = TRUE)
summary(pca)
plot(pca)
```

The values of principal components are stored in pca$x

```{r}
head(pca$x)
```

```{r}
library(ggplot2)
x2 <- data.frame(pca$x, fake = coin$fake)
head(x2)
ggplot(x2, aes(x = PC1, y = PC2, col=as.factor(fake))) +
         geom_point(alpha = 0.5)
```

```{r}
library(plotly)

plot_ly(
  data = x2,
  x = ~PC1,
  y = ~PC2,
  z = ~PC3,
  color = ~as.factor(fake),
  type = "scatter3d",
  mode = "markers",
  marker = list(opacity = 0.5)
)
```


```{r}
pca$rotation
```