---
title: "Notes 2025-11-10"
author: "Blazej Kochanski"
date: "2025-11-10"
output: html_document
---

## The data

```{r warning=FALSE}
library(googlesheets4)
gs4_deauth()
d <- read_sheet("https://docs.google.com/spreadsheets/d/1I1ZJ4I345TfaHvyCCgzMz8pMsfS_J3yi9WcpcrhO6No/edit?usp=sharing", "clean_data")
```

Let's construct a 95-percent CI (confidence interval) for the mean height. We assume the data we collected is a sample from the population of some students. And we ask ourselves what is the average height in the population. 

```{r}
mean(d$height)
test_result<-t.test(d$height, conf.level = 0.95)
print(test_result$conf.int)
```

Compute:

- The confidence interval for the average right hand span in the population (assume 99% confidence). 

- The 95-percent confidence interval for the average movies watching time in hours (`movies`). 

- The 95-percent confidence interval for the proportion of males in the group. 

- The 95-percent confidence interval for the correlation between the height and head size (`head_circ`). 

Some templates from my statistics2 course: 

https://bookdown.org/blazej_kochanski/statistics2/templates-main.html



The confidence interval for the average right hand span in the population (assume 99% confidence). 

```{r}
mean(d$right_hand_span)
test_result<-t.test(d$right_hand_span, conf.level = 0.99)
print(test_result$conf.int)
```

The 95-percent confidence interval for the average movies watching time in hours (`movies`). 

```{r}
mean(d$movies, na.rm=TRUE)
test_result<-t.test(d$movies, conf.level = 0.95)
print(test_result$conf.int)
```

The 95-percent confidence interval for the proportion of males in the group. 

```{r}
(nmales = sum(d$gender=="Male"))
(n = length(d$gender))
#install.packages("binom")
binom::binom.confint(nmales, n, conf.level = 0.95, method="wilson")
```

The 95-percent confidence interval for the correlation between the height and head size (`head_circ`). 

Some templates from my statistics2 course: 

https://bookdown.org/blazej_kochanski/statistics2/templates-main.html


