---
title: "Notes 2025-10-27"
author: "Blazej Kochanski"
date: "2025-10-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Descriptive statistics in R - correlation / association

## The data

```{r}
library(googlesheets4)
gs4_deauth()
d <- read_sheet("https://docs.google.com/spreadsheets/d/1I1ZJ4I345TfaHvyCCgzMz8pMsfS_J3yi9WcpcrhO6No/edit?usp=sharing", "clean_data")
```

## Pearson correlation coefficient

Measures the **linear** correlation between two quantitative (numeric) variables. See [wikipedia](https://en.wikipedia.org/wiki/Pearson_correlation_coefficient)

Example: correlation between right and left hand span. 

```{r}
plot(d$right_hand_span, d$left_hand_span, pch=16, col=rgb(0,0,0,.4))
cor(d$right_hand_span, d$left_hand_span)
```


Correlation between the height and average hand span?

```{r}
d$avg_hand_span = (d$right_hand_span + d$left_hand_span)/2
cor(d$avg_hand_span, d$height)
```

Correlation between the height and average hand span for males? 

```{r}
mask <- (d$gender == "Male") 
plot(d$avg_hand_span[mask], d$height[mask], pch=16, col=rgb(0,0,0,.4))
cor(d$avg_hand_span[mask], d$height[mask])
```

Example [from the history by Galton](https://www.rdocumentation.org/packages/HistData/versions/0.9-3/topics/Galton): correlation between the so-called "midparent" height and an adult child height (both in inches):

```{r}
#install.packages("HistData")
library(HistData)
data(Galton)
plot(jitter(Galton$parent), jitter(Galton$child), 
     pch=16, col=rgb(0,0,0,.4))
cor(Galton$parent, Galton$child)
```